---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# rSPACE Modification for Simulated Occupancy Sampling
## Example Workflow
[rSPACE original creator: Dr. Martha Ellis](https://github.com/mmellis/rSPACE)
Script modifications by: Jordan Heiman
Last updated: 2024-10-18

### Introduction
These script files were originally designed to provide a framework to analyze the statistical power of wildlife occupancy monitoring designs for detecting trends in abundance of a population. These modified scripts expand the capability of the framework to analyze a wider range of sampling designs including those where sampling locations and effort vary during every sampling cycle (i.e. season or year). 

These scripts are not currently setup as a package format but this repository can be cloned and setup as a local R project. This example workflow follows the simulations done by Heiman et al. (in review).

### Library / Functions / Data
#### Library
Begin by installing and loading the packages that are needed for the workflow.

```{r package load, message = T}
source("../2.Functions/00_package_loading.R")
```

#### Functions
Source all the functions that are needed for running the modified rSPACE analysis. As well as sourcing on file with c++ functions that are used for efficiency

```{r function sourcing}
list.files("../2.Functions/",
           pattern = ".R", 
           full.names = TRUE, 
           recursive = FALSE, 
           include.dirs = FALSE) %>% 
  lapply(., source) %>% 
  invisible()

dyn.load("../4.C_Scripts/SPACE.dll")
```






# Data

# Establishes the location of the output folder for saving results to

output_folder \<- here::here("3.Outputs")

## Setup

